Examining spatial equity and accessibility to a public bicycle share program using a balanced floating catchment area approach

### 2021-03-25 Bike-share-spatial-equity/Bike-share-spatial-equity.Rmd
CDH edits
text and code edits

Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016]. In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011]. However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016]. There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020]. However, the relationship between income or education and bike share use is less clear. Stations in disadvantaged communities in Chicago generated most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St. Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019]. Being university educated was also a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011]. On the other hand, members who reside in minority-concentrated and lower socioeconomic status neighbourhoods use the Minneapolis-St Paul bike share more frequently [@wangNeighborhoodSociodemographicCharacteristics2019]. Financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016]. Many studies have found that proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011]. This makes sense given that individuals are more likely to use services or programs that they can easily access. Finally, the potential of bike share programs to increase cycling levels has also been explored recently [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], but more longitudinal evidence is needed to determine their impact on encouraging more cycling and offering more opportunities for physical activity. On the whole, these findings highlight the need for bike share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users, which has been the focus of recent research [@auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020], and to offer more people the option of using sustainable and active transportation.

@@

Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved since its conceptualization and become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020]. As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in usage, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options. It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].

In the context of PBSPs, accessibility refers to the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019]. Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately use of the program, the location and size (e.g., maximum number of bicycles available) of stations matters. Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014]. Kabra et al. [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300m of stations. Furthermore, unlike other public amenities with greater ability to adapt to crowding, like health care services, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical. The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles. Likewise, more people may not opt to use the program if the supply of bicycles available at the nearest station is insufficient to meet demand. The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity. Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts. For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.

Several approaches have been commonly used for measuring accessibility: cumulative opportunities, gravity, and utility-based [@handyMeasuringAccessibilityExploration1997]. Paez et al [-@paezMeasuringAccessibilityPositive2012] provide a recent overview of various formulations and applications of accessibility in transportation research. Briefly, the gravity-based approach involves weighting opportunities, for example the quantity of bicycle share stations, as measured by a function of time, using a negative exponential form [@handyMeasuringAccessibilityExploration1997]. It takes both demand and congestion into account [@paezMeasuringAccessibilityPositive2012]. Floating catchment area (FCA) methods are a type of gravity measure that have been used in healthcare accessibility research but that are applicable to transportation case studies. This approach involves producing flexible catchment areas for populations, recognizing that people may be willing to travel to access particular services [@paezMeasuringAccessibilityPositive2012]. Thus, this model does a good job of considering potential crowding of services using a binary distance function [@paezMeasuringAccessibilityPositive2012]. A recent improvement to this approach was achieved through a simple and intuitive adjustment to the allocation of supply and demand, which addressed the effects of demand and service inflation from the conventional approach [see @paezDemandLevelService2019].

@@

Using balanced floating catchment area methods, a novel approach that has not been used yet in cycling research, we examine accessibility to a public bicycle share program in Hamilton, Ontario. Using disaggregate population-level data, we (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different thresholds of walking to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.

@@

This paper uses the city of Hamilton, located in Ontario, Canada, as a case study. The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015]. Before June 2020, the program was known as SoBi Hamilton. Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021]. The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}]. This represents roughly one fifth of the population in Hamilton, according to the 2016 Canadian Census. The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015]. In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton. An additional 75 bicycles and 15 stations were added to the program, which expanded it to more disadvantaged areas in the core service area [see Figure 2]. The program also offers subsidized memberships to individuals who identify as low income. A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019]. As of June 2020, the bike share program has 900 bikes and 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].

Hamilton Bike Share has conducted one membership survey to date in 2018 [@civicplanSoBiHamiltonMembership2017], and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature]. The majority of respondents live within the core service area and the gender split is typical: 57% of respondents are male and 41% are female. The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups. Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips), and nearly 50% of trips have an average length of 11 to 20 minutes. As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same. This suggests that SoBi Hamilton has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.

@@

Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature. This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019]. In particular, the Two-Step Floating Catchment Area method produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it takes into account individual behaviour and doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019]. This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share. The City of Hamilton has positioned stations between 300 and 600 metres apart, which would lead to boundaries around these buffers, but it would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.

More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that are a result of the overlapping catchment areas produced by the Two-Step Floating Catchment Area [@paezDemandLevelService2019]. This overlap inflates the demand and generates inaccurate or misleading accessibility estimates [@paezDemandLevelService2019]. By adjusting the impedance weights, both supply and demand are proportionally allocated which eliminates the double-counting of population that leads to demand and supply inflation [@paezDemandLevelService2019]. Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of system-wide population and level of service. Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.

In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data. This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences. Applying the balanced floating catchment area approach to examine accessibility to Hamilton Bike Share constitutes a location-based measure with an individual component by stratifying according to median household income [@geursAccessibilityEvaluationLanduse2004], but conducting a further sensitivity analysis by adjusting distance thresholds would introduce additional spatio-temporal constraints to evaluate equitable accessibility.

@@

@@

Following the method first presented by Tobler [-@toblerSmoothPycnophylacticInterpolation1979], we used pycnophylactic interpolation to disaggregate population from each dissemination area to smaller polygons that are 50 by 50 metres. This method involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}]. Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not reallocated or smoothed out to areas where people do not live in Hamilton (i.e., where schools or parks are located). To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton. Next, we extracted these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population [see Figure \ref{fig:interpolated-population}].

@@

Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, therefore the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis. Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010]. We found that accessibility increases with a threshold between two and four minutes, but is then maximized at 5 minutes. At some stations, Hamilton Bike Share will depict a map to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk. Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.

@@

At the minimum threshold, we find that there are 25.2 bicycle racks per person in the original system.  With the addition of equity stations, there are now 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems. Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.

@@

Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 30 minute walk to the service area. We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}]. While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status for dissemination areas across the city not just within the core service area. Instead, we use median total income for each dissemination area in the core service area to examine whether disparities in accessibility exist between income groups for those who live in or near the core service area. We divide income by quantiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%. One of the unique properties of the balanced floating catchment area method is that data can be reaggregated from small to larger polygons while preserving the total population and supply at each station. This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., across dissemination areas instead of 50 by 50 metre population cells).

Panels 1[\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds. Our analysis demonstrates that the extreme threshold of fifteen minutes allows the most people to access a station. We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly [see Table \ref{tab:accessibility-income}]. By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city. This is particularly evident at the minimum and average thresholds of three and five minutes, respectively. The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.

>>>>

Public bicycle share programs (PBSP) have been implemented in over 800 cities worldwide and a great deal has been learned about their typical users [@fishmanBikeshareReviewRecent2016]. In many cities, males use bike share more than females [@breyWantRideMy2017; @nickkarSpatialtemporalGenderLand2019; @reillyGenderDisparitiesNew2020; @wintersWhoAreSuperusers2019] as do younger age cohorts [@breyWantRideMy2017; @buckAreBikeshareUsers2013; @fullerUseNewPublic2011]. However, one study found that bike share users in Washington, DC were more likely to be female [@buckAreBikeshareUsers2013], which suggests that the gender gap among cyclists who use bike share is less disparate than the gap for personal bicycle use [@fishmanBikeshareReviewRecent2016]. There is some evidence that bike share users are less likely to own a car [@buckAreBikeshareUsers2013; @reillyNoncyclistsFrequentCyclists2020]. However, the relationship between income or education and bike share use is less clear. Stations in disadvantaged communities in Chicago generated most of the average annual trips [@qianBikesharingEquityDisadvantaged2020] and individuals from minority or lower socioeconomic status neighbourhoods in Minneapolis-St. Paul used the city's PBSP more [@wangNeighborhoodSociodemographicCharacteristics2019]. Being university educated was also a significant correlate of bike share use in Montreal, Canada [@fullerUseNewPublic2011]. On the other hand, members who reside in minority-concentrated and lower socioeconomic status neighbourhoods use the Minneapolis-St Paul bike share more frequently [@wangNeighborhoodSociodemographicCharacteristics2019]. Financial savings have been found to motivate those on a low income to use bike share [@fishmanBikeshareReviewRecent2016]. Many studies have found that geographic proximity to a bike share station, either living or working near one, is an important determinant of use or having a membership [@fishmanBikeshareReviewRecent2016; @fullerUseNewPublic2011]. This makes sense given that individuals are more likely to use services or programs that they can easily access. Finally, the potential of bike share programs to increase cycling levels has also been explored recently [@hosfordEvaluationImpactPublic2018; @hosfordEvaluatingImpactImplementing2019], but more longitudinal evidence is needed to determine their impact on encouraging more cycling and offering more opportunities for physical activity. On the whole, these findings highlight the need for bike share systems to be more highly accessible for diverse populations in order to increase use beyond the "typical" users, which has been the focus of recent research [@auchinclossDesignBaselineDescription2020; @hullgrassoBikeShareEquity2020; @macarthurAdaptiveBikeShare2020], and to offer more people the option of using sustainable and active transportation.

@@

Accessibility has been applied in both a positive and normative way to inform transportation planning [@paezMeasuringAccessibilityPositive2012], but its utility to this field has evolved over the past century and has increasingly become linked with recent interest in prioritizing local proximity and modes that are suitable for local trips like walking and cycling [@levineCenturyEvolutionAccessibility2020]. As a measure of the ease of reaching potential destinations spread spatially in a given area, accessibility is relevant to PBSPs because it can identify current inequities in usage, as well as guide interventions that increase access for groups that are under-serviced or address gaps in transportation options. It also addresses some of the challenges of other performance measures such as level of service within a transportation network by measuring person-based indicators and exploring differences in use between population groups [@paezMeasuringAccessibilityPositive2012].

Beyond the utility derived from using shared bikes to destinations of value, an important aspect of accessibility associated with PBSPs is the distance an individual must travel to reach a bicycle share station [@kabraBikeShareSystemsAccessibility2020; @wangNewBikeShare2019]. Since the time or distance that members of the population need to reach a bicycle share station decreases the potential of accessing the program and ultimately their use of the program, the location and size (e.g., maximum number of bicycles available) of stations matters. Indeed, distance to bicycle share stations is associated with use [@fullerUseNewPublic2011; @wangNewBikeShare2019] and can be a barrier to using PBSPs [@fishmanBarriersBikesharingAnalysis2014]. Kabra et al. [-@kabraBikeShareSystemsAccessibility2020] found that the majority of bike share usage in Paris comes from areas within 300m of stations. Furthermore, unlike other public amenities with greater ability to adapt to crowding, like health care services, the utility of stations is limited by the number of bicycles that they can hold, which makes crowding effects critical. The program may not necessarily be improved if stations are easy to reach but offer only a small number of bicycles. Likewise, more people may not opt to use the program if the supply of bicycles available at the nearest station is insufficient to meet demand. The location and size of stations is important to increase the utility of this public transportation option for more people, thus achieving vertical equity. Accessibility analyses for PBSPs constitute a positive and evaluation-based approach that also has the potential to inform equity efforts. For instance, Wang and Lindsey [-@wangNewBikeShare2019] investigated whether new or relocated bicycle share stations increased accessibility and use, which offered important insights to improve the performance of the program.

Several approaches have been commonly used for measuring place-based accessibility, including cumulative opportunities, gravity, and utility-based measures [@handyMeasuringAccessibilityExploration1997]. Geurs and Van Wee [-@geursAccessibilityEvaluationLanduse2004] and Paez et al. [-@paezMeasuringAccessibilityPositive2012] provide recent overviews of various formulations and applications of accessibility in transportation research. The common gravity-based approach for example involves weighting destination opportunities, such as the quantity of bicycle share stations, by the time required to reach them from an origin using an impedance function [@handyMeasuringAccessibilityExploration1997; @kwanSpaceTimeIntegral1998]. However, while such measures are suitable for capturing the potential for reaching destinations from a given location, they do not take demand or congestion effects into account.

In contrast, floating catchment area (FCA) methods incorporate information on capacity and demand in calculating accessibility. FCA measures have been widely employed in healthcare accessibility research and are typically calculated across two steps. In the first, a ratio of supply to demand at service locations is calculated, such as the number of beds at a hospital divided by the number of people within the catchment area of the hosptital weighted by the distance involved in reaching the facility. Next, these service level ratios are allocated back to the population centres and summarized as a measure of congested accessibility. Thus, this model does a good job of considering potential crowding or compentition for services. While there have been many methodological innovations in FCA methods **(e.g. [Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120), [Luo & Qi, 2009](https://www.sciencedirect.com/science/article/pii/S1353829209000574?via%3Dihub), [Wan et al., 2012](https://www.tandfonline.com/doi/full/10.1080/13658816.2011.624987), [Delamater, 2013](https://www.sciencedirect.com/science/article/pii/S1353829213001020?via%3Dihub))**, a recent improvement to this approach was achieved through a simple and intuitive balanced FCA approach that addressed the effects of demand and service inflation from previous FCA approaches [see @paezDemandLevelService2019].

@@

Using the balanced floating catchment area method [@paezDemandLevelService2019], a novel approach that has not been used yet in cycling research, with disaggregate population-level data, we examine accessibility to a public bicycle share program in Hamilton, Ontario. We (1) conduct a sensitivity analysis by measuring accessibility and level of service to bike share stations at different thresholds of walking to reach a station: 3 minutes, 5 minutes, 10 minutes, and 15 minutes; (2) explore the contribution of specific stations that were added to Hamilton's PBSP to reducing both horizontal and vertical inequities; and (3) examine whether disparities in accessibility exist according to median household income of dissemination areas within the core service area.

@@

This paper uses the city of Hamilton, located in Ontario, Canada, as a case study. The city launched a public bicycle share program, Hamilton Bike Share, in March 2015 with 115 stations and 750 bicycles [@hamiltonHamiltonBikeShare2015]. Before June 2020, the program was known as Social Bicycles or SoBi Hamilton. Stations are spaced between 300 and 600 metres apart [@scottRouteChoiceBike2021]. The core service area spans 35 km^2 of the city and roughly 138,000 can reach a bike share hub within 30 minutes of walking [see Figure \ref{fig:hamilton-and-sobi-service-area}]. This represents roughly one fifth of the population in Hamilton according to the 2016 Canadian Census. The program was enthusiastically welcomed in the city - within three weeks of launching, 10,000 trips had been made [@hamiltonHamiltonBikeShare2015]. In 2017, Hamilton Bike Share Inc., the non-profit organization that operates the program, initiated an equity program, the Everyone Rides Initiative (ERI), to remove barriers that may prevent individuals from accessing bike share in Hamilton. An additional 75 bicycles and 15 stations were added to the program which expanded it to more disadvantaged areas in the core service area [see Figure 2]. The program also offers subsidized memberships to individuals who identify as low income. A comparable program can be found in Philadelphia [see @caspiBikesharingPhiladelphiaLowerincome2019]. As of June 2020, the bike share program has 900 bikes, 130 stations [see Figure \ref{fig:sobi-stations-in-hamilton}], and over 26,000 active memberships [@hamiltonHamiltonBikeShare2015].

Hamilton Bike Share has conducted one membership survey in 2018 [@civicplanSoBiHamiltonMembership2017] and the findings from 420 members are broadly in line with the trends that were discussed above [see @fishmanBikeshareReviewRecent2016 for a recent review of the literature]. The majority of respondents live within the core service area and the gender split is typical: 57% of respondents are male and 41% are female. The majority of respondents, both male and female, are between 25 and 34 years of age, but the percentage of male respondents is higher in the subsequent age groups. Respondents use bike share for commuting (40% of trips) or errands and meetings (24% of trips) and nearly 50% of trips have an average length of 11 to 20 minutes. As a result of having a bike share membership, 49% of respondents report that they use their private vehicle less often or much less often but 48% report that their private vehicle use has remained about the same. This suggests that SoBi Hamilton has been useful for certain kinds of trips but not all, meaning that some trips continue to require a private vehicle.

@@

Floating catchment area (FCA) methods are an approach commonly used in the healthcare accessibility literature. This approach is more appropriate and informative than calculating the provider-to-population ratio (PPR) which simply divides the level of supply of a service (e.g., the number of bicycle racks at a station) by the population who have access to the service [@paezDemandLevelService2019]. In particular, the Two-Step Floating Catchment Area (2SFCA) method ([Radke & Mu](https://www.tandfonline.com/doi/abs/10.1080/10824000009480538), [Luo & Wang, 2003](https://journals.sagepub.com/doi/10.1068/b29120)) produces flexible catchment areas for populations accessing a service rather than using rigid boundaries like PPR, which provides more useful information because it doesn't assume that people stay within the pre-defined boundaries [@paezDemandLevelService2019]. This is an important property that supports our rationale for applying this method to measure accessibility to Hamilton Bike Share. The City of Hamilton has positioned stations between 300 and 600 metres apart, which would lead to boundaries around these buffers, but it would be reasonable to assume that people are willing to walk beyond this threshold to access other stations if the ones nearest them have no supply of bicycles.

More recently, the _balanced_ floating catchment area approach was developed to address issues with demand and supply inflation that result from the overlapping catchment areas produced by past FCA methods [@paezDemandLevelService2019]. Briefly, the overlapping of catchment areas leads to inflation of population totals and deflation of service levels across a study area and generates inaccurate or misleading accessibility estimates. In contrast, Paez et al. [-@paezDemandLevelService2019] adjust the impedance weights so that both supply and demand are proportionally allocated which eliminates the over-counting of population that leads to distortions in demand and supply. Other benefits of this adjusted method include consideration of competition which can occur when catchment areas overlap, as well as the preservation of system-wide population and level of service. Balanced floating catchment area methods have been used to explore accessibility to health care providers [@paezDemandLevelService2019] and COVID-19 health care services [@pereiraGeographicAccessCOVID192021], but have not yet been used in the cycling literature to explore issues of accessibility.

@@

For this paper, we employ a hybrid place-based and person-based approach to calculating accessibility using disaggregate population data. In their review of accessibility measures, Geurs and van Wee [-@geursAccessibilityEvaluationLanduse2004] highlight the need for greater inclusion of individual spatio-temporal constraints but acknowledge the challenges of acquiring and analyzing person-based data. This comes after Kwan's [-@kwanSpaceTimeIntegral1998] work to show that space-time measures are more capable of capturing interpersonal differences, especially the effect of space-time constraints on individual behaviour, and are more helpful for unraveling gender/ethnic differences. Applying the balanced floating catchment area approach to examine accessibility to Hamilton Bike Share constitutes a location or place-based measure with an individual component by stratifying according to median household income [@geursAccessibilityEvaluationLanduse2004], but conducting a further sensitivity analysis by adjusting distance thresholds would introduce additional spatio-temporal constraints to evaluate equitable accessibility.

@@

Following the method first presented by Tobler [-@toblerSmoothPycnophylacticInterpolation1979], we used pycnophylactic interpolation to disaggregate population from each dissemination area to smaller polygons that are 50 by 50 metres. This method involves smoothing out the population from each dissemination area while preserving total volume [see Figure \ref{fig:da-population}]. Since pycnophylactic interpolation occurs at such a micro scale, we had to ensure that population numbers were not reallocated or smoothed out to areas where people do not live in Hamilton (i.e., where schools or parks are located). To do so, we retrieved shapefiles for various geographic features [see Table \ref{tab:data-features}] from Open Hamilton. Next, we removed these features from the PBSP core service area and used pycnophylactic interpolation to disaggregate and reallocate population within the remaining area [see Figure \ref{fig:interpolated-population}].

@@

Consensus regarding the distance that individuals are willing to walk to access a PBSP is lacking, therefore the literature on walking behaviour was consulted to determine the thresholds for the sensitivity analysis. Previous studies have found that living within 250 metres [@fullerUseNewPublic2011] and 300 metres [@kabraBikeShareSystemsAccessibility2020] is correlated with bike share use, while other research has found that walking trips are less than 600 metres and rarely more than 1200 metres [@millwardActivetransportWalkingBehavior2013] or a median distance of 650 metres [@larsenQuarterMileReexamining2010]. In Hamilton, Hamilton Bike Share will depict a map at some stations to show the user the locations of the other nearest stations within a five minute walk, which suggests that this is an average distance that people are anticipated to walk. In the present case, we found that congested accessibility calculated using the balanced FCA approach increases with a threshold between two and four minutes, but is then maximized at 5 minutes. Accessibility decreases significantly after eight minutes, which is intuitive given that demand on a limited supply increases as more people can reach each bike share station.

@@

At the minimum threshold, we find that there are 25.2 bicycle racks per person in the original system.  The addition of equity stations increases this ratio slightly to 25.4 bicycles per person. Figure \ref{fig:figure-6} presents a comparison of accessibility between the systems for the population cells. Accessibility is fairly uniform, with the exception of two small areas near the university and the waterfront park where accessibility is slightly higher.

@@

Approximately 138,000 people live within Hamilton Bike Share's core service area or within a 30 minute walk to the service area. We find that over 118,000 people can access a bicycle share station within a 15 minute walk, which represents roughly 85% of the total population in the core service area [see Table \ref{tab:accessibility-income}]. While previous research found that neighbourhoods with more disadvantage are better serviced by Hamilton Bike Share [@hosfordWhoArePublic2018], the authors used the Pampalon Deprivation Index from 2011 as a measure of socioeconomic status for dissemination areas across the city not just within the core service area. Instead, we use median total income for each dissemination area in the core service area to examine whether disparities in accessibility exist between income groups for those who live in or near the core service area. We divide income by quintiles: bottom 20%, second 20%, third 20%, fourth 20%, and top 20%. One of the unique properties of the balanced floating catchment area method is that data can be re-aggregated from small to larger polygons while preserving the total population and supply at each station. This avoids demand and supply inflation, and also enables us to present findings in a way that is easier to interpret (i.e., across dissemination areas instead of 50 by 50 metre population cells).

Panels 1 [\ref{fig:figure-bi-map-threshold-3}], 2 [\ref{fig:figure-bi-map-threshold-5}], 3 [\ref{fig:figure-bi-map-threshold-10}], and 4 [\ref{fig:figure-bi-map-threshold-15}] depict bivariate choropleth maps with the combined spatial distribution of accessibility and median total income per household for the different thresholds. Our analysis demonstrates that the extreme threshold of fifteen minutes allows the most people to access a station. We find that stations added to Hamilton's public bicycle share program to increase equity for disadvantaged neighbourhoods achieved their goal by increasing accessibility, albeit only modestly [see Table \ref{tab:accessibility-income}]. By implementing equity stations in more areas with lower median total income in Hamilton, the PBSP has achieved greater horizontal equity by extending the spatial distribution of bicycles across the city. This is particularly evident at the minimum and average thresholds of three and five minutes, respectively. The largest gains were made for dissemination areas in the second 20%, where an additional 3,073 and 5,395 people could reach a bicycle share station within three and five minutes walk, respectively, after the addition of the equity station.
